Semantic Preserving Bijective Mappings for Expressions involving Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-30 main.tex
Editing all HSC comments (close #30)

In order to clarify the notion of \textit{equivalence} (or lack thereof) in our context of translations,
we introduce the terms \textit{appropriate} and \textit{inappropriate} translations. 
%{\bf [HSC: I don't understand how {\it equivalent} translations has anything at 
%all to do with {\it appropriate} or {\it inapppropriate} translations. Please clarify.]}.
We consider a translation to be {\it appropriate}, when a numerical evaluation returns the 
same values in both concepts up to a numerical precision $|\epsilon|\ll 1$, for all possible 
points in specified domains for the functions.
A translation is considered as {\it inappropriate}, when it is not {\it appropriate}.

@@

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects {\bf [HSC: which two projects? Please clarify.]}. The goal of \sTeX{} is to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro functionality developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive number of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is often complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not usually support.

@@

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care {\bf [HSC: ``takes care'' is too informal. Use a different phrase or word.]} of each translation and provides details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions $w_1$, $w_2$, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
\mathscr{W}\{ w_1(z), w_2(z) \} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{}, it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
This translation is unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$, $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the variable of differentiation $z$. We solved this issue by creating a new macro \verb|\Wron|, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and further updates are sometimes
necessary in order to further encapsulate critical semantic information {\bf [HSC: I changed this sentence. Check to make sure that it is still correct.]}.

\subsection{Potentially Ambiguous Expressions}

{\bf [HSC: This section needs to start out with a short overview paragraph describing the issue instead of an example. Also, I changed the title of the section.]}

A similar problem is multiplications since they are rarely explicitly marked in \LaTeX{} expressions, e.g., scientists using whitespace to indicate multiplications rather than using \verb|\cdot| or similar symbols. For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication symbol (this macro will not be rendered). Since this macro is newly introduced by contributers of the \gls*{drmf} team, and automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} use this macro. Therefore, the translator has some simple 
rules for performing translations without explicitly marking multiplication translations with \verb|\idot|.

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' gives the impression that $4$ indexing `$b$'. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons:~(1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we have used rules as follows {\bf [HSC: I changed the wording here, definition$\to$rules. Is this ok? Otherwise further clarify.]}

@@

In general, the translator is designed to find a work-around for disambiguating expressions. 
If there is no way to solve a potential ambiguity with defined rules, then we stop the translation process
{\bf [HSC: I changed the wording here also. Is this ok?]}.

@@

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that define translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder used to describe {\bf [HSC: I used describe instead of define. Is this ok? Otherwise use a different more descriptive term.]} the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

\section{Forward Translations}\label{sec:forward-translation}
As a pre-processing step, we use the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.} for parsing semantic \LaTeX{} expressions. The \gls*{pom}-Tagger is defined by a context-free grammar in \gls*{bnf} and is an \texttt{LL}-Parser, i.e., it parses the input from \textbf{L}eft to right and assigns the \textbf{L}eftmost (first applicable) derivation rule defined by the grammar to an expression. In other words, the \gls*{pom}-Tagger scans the input for \textit{terms} and groups them into subexpressions if suitable, where \textit{terms} are non-terminal symbols in the context of \gls*{bnf}. If a node in the generated parse tree matches a pre-defined symbol, it will be tagged by mata {\bf [HSC: mata? If this is correct, please define, otherwise clarify.]} information that are defined in manually cultivated lexicon files {\bf [HSC: please reword this sentence because it reads poorly.]}.

@@

\Cref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from~\Cref{tab:JacobiP-usecase}. Because of the differences to {\bf [HSC: to? does not sound right? Pleased use a better word. between? Also, clean up the sentence. It does not read so well.]} expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}. The \gls*{teo} is a list containing already translated subexpressions.

@@

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$, etc.), while functional notations are conveniently used for any kind of functions ($\sin$, $\cos$, etc.). Sometimes the same symbol is used in different notations to distinguish different meanings. For example, the `$-$' as a unary operator is used in prefix notation to indicate the negative value of its operand, such as in `$-2$'. Of course, `$-$' can also be the binary operator for subtraction, which is commonly used in infix notation. One example is the postfix notation used with factorials, such as for `$2!$'
{\bf [HSC: I changed the wording slightly here. Is this ok?]}.

Since it is more convenient to parse expressions using uniform notations, most programming 
languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix 
the notations in one expression {\bf [HSC: I reworded this sentence. Is it ok?]}.
However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, there are two new problems with this approach.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

@@

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example should not be misinterpreted since this notation is the standard interpretation in science for the double factorial. We wrote the second interpretation with parentheses point it out more precisely {\bf [HSC: please rewrite this sentence.]}. However, surprisingly, \Maple{} computes the first interpretation (the factorial of the factorial of $n$) again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1:~& \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2:~& $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples using the factorial and double factorial symbols. One expression in a text format can potentially be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}
\vspace*{-0.5cm}

@@

\subsection{Subtranslators}\label{sec:subtranslators}

{\bf [HSC: Is it `The \verb|SequenceTranslator|' or is it `A \verb|SequenceTranslator|'? Please choose. You cannot use both. Once you have decided implement throughout the paper.]}

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on), the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the 
\verb|SequenceTranslator| 
also checks balancing of parentheses in expressions. An expression such as `$(a]$' produces a mismatched parentheses error. On the other hand, this is a problem for real interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

@@

	\State \textit{pattern} = \textit{info}.getTranslationPattern();
	\State \textit{translatedMacro} = \textit{pattern}.fillPlaceHolders(\textit{argList});\label{line:fill_pattern}
	\If{\textit{power} is not \NULL}\label{line:shifted_exp}
		\State \textit{translatedMacro}.add(\textit{power});
	\EndIf
	\State \Return\ \textit{translatedMacro};
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\vspace*{-0.2cm}

{\bf [HSC: This is a new paragraph, but it does not have standard vertical space above it. I do not know why this is. Can you fix this?]} The translation process for the \Macro s is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describe the different kinds of the following siblings after a semantic macro node. Those can be:
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were 
passed (Line~\ref{line:trans_vars}).
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to access the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations. Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to obtain the syntax tree of the input\footnote{A license of \Maple{} is mandatory to perform backward translations. Our translator uses the version \Maple{} 2016 {\bf [HSC: I reworded this. Is it ok? Otherwise fix.]}.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes 
and is also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

\section{Evaluation}\label{sec:evaluation}
We implemented three approaches to evaluate whether a translation was \textit{appropriate} or \textit{inappropriate}.
\begin{enumerate}
\item \textbf{Round Trip Tests}:~translates expressions back and forth and analyzes the changes.
\item\label{approach-2} \textbf{Function Relation Tests (Symbolical)}:~translates mathematically proven equivalent expressions from one system to a \gls*{cas} and evaluates whether the relation remains valid via symbolical equivalence checks.
\item \textbf{Numerical Tests}:~takes the same equations from Approach~\ref{approach-2} but evaluates them on specific numerical values to test equivalence.
\end{enumerate}

@@

With this technique {\bf [HSC: what technique? A technique has not been introduced yet in this section. Please describe this better, or be more specific.]}, we can detect translation errors such as in \eqref{eq:wrong-trans-1} and \eqref{eq:wrong-trans-2}. Consider the \gls*{dlmf} equation for the sine and cosine function~\parencite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin \left(u+v\right) = \sin{u}\cos{v} + \cos{u}\sin{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Then
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos\left(u+v\right) = \cos{u}\sin{v} + \sin{u}\cos{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude that there was an error 
during the translation process.

>>>>

In order to clarify the notion of \textit{equivalence} (or lack thereof) in our context of translations,
we introduce the terms \textit{appropriate} and \textit{inappropriate} translations. 
We consider a translation to be {\it appropriate}, when a numerical evaluation returns the 
same values in both concepts up to a numerical precision $|\epsilon|\ll 1$, for all possible 
points in specified domains for the functions.
A translation is considered as {\it inappropriate}, when it is not {\it appropriate}.

@@

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using  the semantic macros introduced by Miller. The goal of \sTeX{} is to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro functionality developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive number of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is often complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not usually support.

@@

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A particular logging function stores each translation and provides details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions $w_1$, $w_2$, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
\mathscr{W}\{ w_1(z), w_2(z) \} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{}, it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
This translation is unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$, $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the variable of differentiation $z$. We solved this issue by creating a new macro \verb|\Wron|, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and further updates are sometimes
necessary in order to further encapsulate critical semantic information.

\subsection{Potentially Ambiguous Expressions}
Since the \Macro s aims to cover an extensive set of special functions, orthogonal polynomials, and mathematical constants, they does not contain specific macros for other mathematical objects. However, also mathematical expression without functions, polynomials and mathematical constants can be ambigious. As an example, multiplications are rarely explicitly marked in \LaTeX{} expressions, e.g., scientists using whitespace to indicate multiplications rather than using \verb|\cdot| or similar symbols. But whitespaces can also be used to improve the readability and not to represent a multiplication.

For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication symbol (this macro will not be rendered). Since this macro is newly introduced by contributers of the \gls*{drmf} team, and automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} use this macro. Therefore, the translator has some simple 
rules for performing translations without explicitly marking multiplication translations with \verb|\idot|.

@@

Another more questionable translation decision addresses alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' gives the impression that $4$ indexing `$b$'. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons:~(1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we have used rules as follows

@@

In general, the translator is designed to find a work-around for disambiguating expressions. 
If there is no way to solve a potential ambiguity with defined rules, then we stop the translation process.

@@

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that define translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder to determine the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

\section{Forward Translations}\label{sec:forward-translation}
As a pre-processing step, we use the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.} for parsing semantic \LaTeX{} expressions. The \gls*{pom}-Tagger is defined by a context-free grammar in \gls*{bnf} and is an \texttt{LL}-Parser, i.e., it parses the input from \textbf{L}eft to right and assigns the \textbf{L}eftmost (first applicable) derivation rule defined by the grammar to an expression. In other words, the \gls*{pom}-Tagger scans the input for \textit{terms} and groups them into subexpressions if suitable, where \textit{terms} are non-terminal symbols in the context of \gls*{bnf}. A node in the generated parse tree will be tagged by meta information if the node matches defined symbols. The meta information is stored in lexicon files. Those lexicon files were manually cultivated for the \gls*{pom}-Tagger.

@@

\Cref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from~\Cref{tab:JacobiP-usecase}. Because of the differences between expression trees and \gls*{pom-pt}, it can be difficult to generate a string representation after a successful translation process. It is especially difficult to determine necessary and unnecessary parentheses when we analyse the \gls*{pom-pt}. Therefore we create the \gls*{teo}. The \gls*{teo} is a list containing already translated subexpressions.

@@

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$, etc.), while functional notations are conveniently used for any kind of functions ($\sin$, $\cos$, etc.), and the postfix notation is often common for unary operators ($2!$, $-2$, etc.). Sometimes the same symbol is used in different notations to distinguish different meanings. For example, the `$-$' as a unary operator is used in prefix notation to indicate the negative value of its operand, such as in `$-2$'. Of course, `$-$' can also be the binary operator for subtraction, which is commonly used in infix notation. 
% [AGP: I moved the 2! example to the first sentence where I explain all kinds of notations

Since it is more convenient to parse expressions using uniform notations, most programming 
languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix 
the notations in one expression.
However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, there are two new problems with this approach.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

@@

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example should not be misinterpreted since this notation is the standard interpretation in science for the double factorial. We wrote the second interpretation with parentheses for pointing out that we mean the double factorial in this case. However, surprisingly, \Maple{} computes the first interpretation (the factorial of the factorial of $n$) again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1:~& \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2:~& $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples using the factorial and double factorial symbols. One expression in a text format can potentially be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}
\vspace*{-0.5cm}

@@

\subsection{Subtranslators}\label{sec:subtranslators}
The \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on), the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the 
\verb|SequenceTranslator| 
also checks balancing of parentheses in expressions. An expression such as `$(a]$' produces a mismatched parentheses error. On the other hand, this is a problem for real interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

@@

	\State \textit{pattern} = \textit{info}.getTranslationPattern();
	\State \textit{translatedMacro} = \textit{pattern}.fillPlaceHolders(\textit{argList});\label{line:fill_pattern}
	\If{\textit{power} is not \NULL}\label{line:shifted_exp}
		\State \textit{translatedMacro}.add(\textit{power});
	\EndIf
	\State \Return\ \textit{translatedMacro};
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

The translation process for the \Macro s is complex, so there is a special class, the \verb|MacroTranslator|, that handles those nodes in the \gls*{pom-pt}. Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has, in Line~\ref{line:get_info}. It also processes the following siblings to translate the arguments. The \verb|MacroTranslator| will be called in Line~\ref{line:transLeaf} in Algorithm~\ref{alg:translation}, since the macro is a leaf node in the \gls*{pom-pt}. The following cases describe the different kinds of the following siblings after a semantic macro node. Those can be:
\begin{itemize}
\item an exponent, such as for `\verb|^2|' right after the macro node (Line~\ref{line:next_caret});
\item an optional parameter in square brackets right after the macro node or after an exponent (Line~\ref{line:next_optional});
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above and no `$@$' symbols were passed (Line~\ref{line:trans_paras});
\item `$@$' symbols (Line~\ref{line:skip_ats}); or
\item a variable in curly brackets (a \textit{sequence} node) after the `$@$' symbols were 
passed (Line~\ref{line:trans_vars}).
\end{itemize}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to access the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations. Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to obtain the syntax tree of the input\footnote{A license of \Maple{} is mandatory to perform backward translations. Our translator uses the version \Maple{} 2016.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes 
and is also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

\section{Evaluation}\label{sec:evaluation}
We implemented three approaches to evaluate whether a translation was \textit{appropriate} or \textit{inappropriate}.
\begin{enumerate}
\item \textbf{Round Trip Tests}:~translates expressions back and forth and analyzes the changes.
\item\label{approach-2} \textbf{Function Relation Tests (Symbolical)}:~translates mathematically proven equivalent expressions from one system to a \gls*{cas} and evaluates whether the relation remains valid via symbolical equivalence checks.
\item \textbf{Numerical Tests}:~takes the same equations from Approach~\ref{approach-2} but evaluates them on specific numerical values to test whether the translation was {\it appropriate}.
\end{enumerate}

@@

With this idea, we can detect {\it inappropriate} translations such as in \eqref{eq:wrong-trans-1} and \eqref{eq:wrong-trans-2}. Consider the \gls*{dlmf} equation for the sine and cosine function~\parencite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin \left(u+v\right) = \sin{u}\cos{v} + \cos{u}\sin{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Then
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos\left(u+v\right) = \cos{u}\sin{v} + \sin{u}\cos{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude that there was an error 
during the translation process and our defined translations were {\it inappropriate}.

However, there are two essential problems with this approach. Testing whether expressions are {\it appropriate} representations of each other is a challenging task for \gls*{cas} and they often have difficulties testing simple equations symbolically. For example, 
consider
\parencite[(4.35.34)]{NIST:DLMF}
\begin{equation*}
\sinh \left( x+\iunit y \right) = \sinh x \cos y + \iunit \cosh x \sin y,
\end{equation*}
as a difference of the left- and right-hand sides cannot be simplified to zero by default. Furthermore, this approach only checks forward translations because there is no way to automatically check whether two \LaTeX{} expressions are {\it appropriate} or {\it inappropriate} representations of each other (again this could become feasible with our translator). We use \Maple's \texttt{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \texttt{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. The simplification function is the most powerful function to check whether two expressions are {\it appropriate} representations in \Maple. However, there are several cases where simplification fails. Because of implementation details, there are some techniques that help \Maple{} to find possible simplifications. For example, we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an {\it appropriate} representation using their exponential representations, namely
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions, we are able to improve the simplification process in \Maple. However, the limitations of the \texttt{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ is the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains a $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one should instead use the analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the previous checks for {\it appropriate} representations in \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.
